Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation

### 2013-02-28 manuscript/manuscript.tex
everything is now up-to-speed with auto R output

Figure \ref{chr2roc} shows an ROC plot of the performance of each algorithm on the simulated data set. All approaches show excellent specificity at high thresholds in this simulation. Cloudbreak provides the greatest specificity at higher levels of sensitivity, followed by DELLY. Table \ref{chr2preds} shows the total number of simulated deletions found by each tool when using a threshold that gives an FDR of 0.1 based on the ROC curve, representing cutoffs of 2.1 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 12 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and 33 supporting read pairs for Pindel. Cloudbreak shows excellent sensitivity at detecting deletions from each size class, and contributes a number of exclusive true predictions in both small and medium size classes.

\begin{table}[t]
\begin{center}
\begin{tabular}{rrrrrr}
  \hline
 & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number & 224 &  84 & 82 &  31 & 26\\ 
  \hline
  Cloudbreak  & 51 (10)  & 51 (\textbf{3}) &  \textbf{55} (\textbf{5}) & \textbf{12} (\textbf{4}) & \textbf{15} (\textbf{1}) \\ 
  Breakdancer & \textbf{52} (14)  & 49 (2) &  49 (0) & 7 (0) & 14 (0) \\ 
  GASVPro     & 31 (4)  & 25 (0) &  23 (0) & 2 (0) & 6 (0) \\ 
  DELLY       & 22 (2)  & \textbf{55} (2) &  40 (0) & 8 (0) & 12 (0) \\ 
  Pindel      & 29 (\textbf{16})  & 11 (0) &  21 (0) & 1 (0) & 8 (0)\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter deletions found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses). The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
\label{chr2preds}
\end{table}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.75\textwidth]{NA18507_ROC.pdf}
\caption{ROC curve of deletion prediction performance on the 37X NA18507 sample, tested against the combined gold standard deletion set taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.}
\label{NA18507roc}
\end{figure}

@@

We then tested the predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X. The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those cutoffs, is shown in Table \ref{NA18507preds}. Cloudbreak scores well in every category: it identifies more small deletions than any tool except for Pindel, which has low specificity at this cutoff, and contributes more exclusive true positives in larger size classes than any other tool.

% latex table generated in R 2.15.2 by xtable 1.7-0 package
% Tue Jan  1 22:36:02 2013
\begin{table}[t]
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
 & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
Total Number & & & 7,466 & 235 & 218 & 110 & 375 \\
  \hline
Cloudbreak & 0.098 & \textbf{0.115} & \textbf{423} (163) & 128 (\textbf{9}) & 158 (\textbf{8}) &  70 (\textbf{4}) & 186 (12) \\ 
Breakdancer & 0.097 & 0.111 & 258 (42) & \textbf{132} (8) & \textbf{167} (2) & \textbf{92} (1) & \textbf{288} (\textbf{14}) \\
  GASVPro & \textbf{0.132} & 0.040 & 102 (11) &  37 (2) & 77 (0) &  26 (0) & 93 (0) \\ 
  DELLY & 0.081 & 0.091 & 143 (10) & 125 (8) & 158 (1) &  83 (1) & 256 (3) \\ 
  Pindel & 0.013 & 0.079 & 406 (\textbf{290}) & 40 (0) & 90 (0) &  37 (0) & 88 (0) \\ 
   \hline
\end{tabular}
\end{center}
\caption{The precision and recall with respect to the gold standard set of each tool on the NA18507 data, as well as the number of deletions in each size class found. The same cutoffs were used as for the simulated data as in Table \ref{chr2preds}, adjusted for the difference in coverage from 30X to 37X.}
\label{NA18507preds}
\end{table}

@@

We analyzed the predictions made by Cloudbreak to determine whether any particular class of deletion event was contributing to its performance. We found that Cloudbreak shows particularly strong performance in detecting deletions that overlap RepeatMasker-annotated elements (Table \ref{repmaskpreds}) in both the simulated and NA18507 data sets, especially compared to the RP-based methods. In the simulated data set, Cloudbreak correctly identifies 9\% more repeat-overlapping deletions than the nearest competitor. Cloudbreak identifies more deletions in repeats than any other method, and in the NA18507 data set finds 3.75 times more exclusive hits than the other RP-based methods. Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes greatly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible.

\begin{table}[b]
\begin{center}
\begin{tabular}{rrr|rr}
 & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\hline
 &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
 Total Number & 120 & 327 & 553 & 7851 \\ 
  \hline
  Cloudbreak  & \textbf{29} (4) & \textbf{155} (\textbf{19}) & \textbf{204} (\textbf{38}) & \textbf{761} (158) \\ 
  Breakdancer & \textbf{29} (\textbf{6}) & 142 (10) & 186 (25) & 751 (42) \\
  GASVPro     & 15 (2) & 72 (2) & 71 (5) & 264 (8) \\
  DELLY       & 21 (2) & 116 (2) & 147 (12) & 618 (11) \\
  Pindel      & 8 (4) & 62 (12) & 98 (26) & 563 (\textbf{264}) \\ 
   \hline
\end{tabular}
\end{center}
\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.}
\label{repmaskpreds}
\end{table}

>>>>

Figure \ref{chr2roc} shows an ROC plot of the performance of each algorithm on the simulated data set. All approaches show excellent specificity at high thresholds in this simulation. Cloudbreak provides the greatest specificity at higher levels of sensitivity, followed by DELLY. Table \ref{chr2DeletionPreds} shows the total number of simulated deletions found by each tool when using a threshold that gives an FDR of 0.1 based on the ROC curve, representing cutoffs of 2.1 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 12 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and 33 supporting read pairs for Pindel. Cloudbreak shows excellent sensitivity at detecting deletions from each size class, and contributes a number of exclusive true predictions in both small and medium size classes.

\begin{table}[t]
\begin{center}
\begin{tabular}{rrrrrr}
  \hline
 & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number & 224 &  84 & 82 &  31 & 26\\ 
  \hline
  Cloudbreak  &   51 (  10)  &   51 (\textbf{   3}) &  \textbf{  55} (\textbf{   5}) & \textbf{  12} (\textbf{   4}) & \textbf{  15} (\textbf{   1}) \\ 
  Breakdancer & \textbf{  52} (  14)  &   49 (   2) &    49 (0) &    7 (0) &   14 (0) \\ 
  GASVPro     &   31 (   4)  &   25 (0) &    23 (0) &    2 (0) &    6 (0) \\ 
  DELLY       &   22 (   2)  & \textbf{  55} (   2) &    40 (0) &    8 (0) &   12 (0) \\ 
  Pindel      &   29 (\textbf{  16})  &   11 (0) &    21 (0) &    1 (0) &    8 (0)\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses). The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
\label{chr2DeletionPreds}
\end{table}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.75\textwidth]{NA18507_DELS_ROC.pdf}
\caption{ROC curve of deletion prediction performance on the 37X NA18507 sample, tested against the combined gold standard deletion set taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.}
\label{NA18507roc}
\end{figure}

@@

We then tested the predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X. The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those cutoffs, is shown in Table \ref{NA18507DeletionPreds}. Cloudbreak scores well in every category: it identifies more small deletions than any tool except for Pindel, which has low specificity at this cutoff, and contributes more exclusive true positives in larger size classes than any other tool.

\begin{table}[t]
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
 & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
Total Number & & & 7,466 & 235 & 218 & 110 & 375 \\
  \hline
Cloudbreak & 0.0978 & \textbf{0.115} & \textbf{ 423} ( 163)  &  128 (\textbf{   9}) &   158 (\textbf{   8}) &   70 (\textbf{   4}) &  186 (  12) \\ 
Breakdancer & 0.119 & 0.111 &  258 (  42)  & \textbf{ 132} (   8) &  \textbf{ 167} (   2) & \textbf{  92} (   1) & \textbf{ 288} (\textbf{  14}) \\  
  GASVPro & \textbf{0.132} & 0.0399 &  102 (  11)  &   37 (   2) &    77 (0) &   26 (0) &   93 (0) \\ 
  DELLY & 0.0813 & 0.091 &  143 (  10)  &  125 (   8) &   158 (   1) &   83 (   1) &  256 (   3) \\ 
  Pindel & 0.0134 & 0.0787 &  406 (\textbf{ 290})  &   40 (0) &    90 (0) &   37 (0) &   88 (0) \\ 
   \hline
\end{tabular}
\end{center}
\caption{The precision and recall with respect to the gold standard set of each tool on the NA18507 data, as well as the number of deletions in each size class found. The same cutoffs were used as for the simulated data as in Table \ref{chr2DeletionPreds}, adjusted for the difference in coverage from 30X to 37X.}
\label{NA18507DeletionPreds}
\end{table}

@@

We analyzed the predictions made by Cloudbreak to determine whether any particular class of deletion event was contributing to its performance. We found that Cloudbreak shows particularly strong performance in detecting deletions that overlap RepeatMasker-annotated elements (Table \ref{deletionRepmaskpreds}) in both the simulated and NA18507 data sets, especially compared to the RP-based methods. In the simulated data set, Cloudbreak correctly identifies 9\% more repeat-overlapping deletions than the nearest competitor. Cloudbreak identifies more deletions in repeats than any other method, and in the NA18507 data set finds 3.75 times more exclusive hits than the other RP-based methods. Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes greatly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible.

\begin{table}[b]
\begin{center}
\begin{tabular}{rrr|rr}
 & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\hline
 &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
 Total Number & 120 & 327 & 553 & 7851 \\ 
  \hline
  Cloudbreak  & \textbf{  29} (   4) & \textbf{ 155} (\textbf{  19}) & \textbf{ 204} (\textbf{  38}) & \textbf{ 761} ( 158) \\ 
  Breakdancer & \textbf{  29} (\textbf{   6}) &  142 (  10) &  186 (  25) &  751 (  42) \\
  GASVPro     &   15 (   2) &   72 (   2) &   71 (   5) &  264 (   8) \\
  DELLY       &   21 (   2) &  116 (   2) &  147 (  12) &  618 (  11) \\
  Pindel      &    8 (   4) &   62 (  12) &   98 (  26) &  563 (\textbf{ 264}) \\ 
   \hline
\end{tabular}
\end{center}
\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.}
\label{deletionRepmaskpreds}
\end{table}
